\documentclass{article}

\usepackage{technical_notes_1_20250424}

\makeindex

\begin{document}

\tableofcontents

\section{Lecture 1}
\label{sec:lecture1}

Today, we reviewed:
\begin{enumerate}
\item Homology.
\item How to construct cohomology.
\item Made a statement about a sequence splitting.
\item The sequence contains $\ker \delta_n$, which we don't understand.
\end{enumerate}

Understanding this $\ker \delta_n$ will be the topic of the next lecture. For more details, see my handwritten notes.

Recall that \textbf{homology}\index{homology} is represented as taking a space $X$, making a chain complex $C_i$ out of it, and then by a purely algebraic manipulation creating a homology $H_i$. For a cochain complex, we dualize $C_i$, then take a cohomology $H^i$.

\begin{equation}
\begin{tikzcd}[column sep=5.0em]
  \text{Space } X \arrow[r, rightsquigarrow] & \text{chain complex } C_i \arrow[r, rightsquigarrow, "\text{purely algebraic}"] \arrow[d, "\text{dualize}", rightsquigarrow] & \text{homology} H_i \\
  & \text{cochain complex } C_i \ar[r, rightsquigarrow] & \text{cohomology } H^{i}
\end{tikzcd}
\end{equation}

Suppose now that we have a chain complex $C_{i}$ of free Abelian groups:
\begin{equation}
  \dots C_{n + 1} \xrightarrow{\partial} C_{n} \rightarrow C_{n - 1} \rightarrow \dots.
\end{equation}
Take an abelian group $G$ (feel free to think about $G = \Z$) and consider the dual
\begin{equation}
  (C_n)^{*} = \text{Hom}(C_n, G) \text{, an additive group.}
\end{equation}

We have no natural map from $C_n^{*} \rightarrow C_{n - 1}^{*}$, but we do have one from 
\begin{align}
\label{eq:1}
  C_n^{*} &\rightarrow C_{n + 1}^{*} \\
  \partial^{*} : \text{Hom}(C_n, G) &\rightarrow \text{Hom}(C_{n+1}, G) \\
  \varphi &\mapsto \varphi \circ \partial
\end{align}

giving us 
\begin{align}
\label{eq:2}
  \dots \rightarrow C_{n - 1}^{*} \rightarrow C_n^{*} \xrightarrow{\delta} C_{n + 1}^{*} \rightarrow C_{n+2}^{*} \rightarrow \dots
\end{align}

Note that we use the sign convention
\begin{equation}
  \delta : C_n^{*} \rightarrow C_{n+1}^{*} \text{ in } (-1)^{n + 1}\partial^{*}
\end{equation}
which is \textbf{\textit{not}} used in Hatcher \& Bredon.

\begin{Definition}{Cochain Complex}
. A \textbf{cochain complex}\index{cochain complex} $C^{\bullet}$ is a collection of Abelian groups $C^n$ with codifferentials $\delta_n$ such that $\delta_{n + 1}\delta_n = 0$.
\end{Definition}

Thus,
\begin{equation}
  \mathcal{H}om(C_{\bullet}, G) = \text{Hom}(C_i, G) \text{ with differential } \delta = (-1)^{n + 1}\partial^{*}
\end{equation}
is a cochain complex, as defined above. Be cautious of $\text{Hom}$ vs. $\mathcal{H}om$.

\begin{Definition}{Cohomology}
. For a cochain complex $C^{\bullet}$, define \textbf{cohomology}\index{cohomology} as
\begin{equation}
  H^i(C^{\bullet}) = \frac{\ker \delta_i}{\text{im} \delta_{i-1}}.
\end{equation}
\end{Definition}

If $C_{\bullet}$ is a chain complex, then any choice of abelian group $G$ gives a cochain complex
\begin{equation}
  \mathcal{H}om(C_{\bullet}, G),
\end{equation}
hence a cohomology and thus gohomology groups, denoted as $H^{*}(C_{\bullet}, G)$.

If $C_{\bullet}$ is the singular cochain complex of a space $X$, then put
\begin{equation}
  H^{*}(X', G) = H^{*}(\mathcal{H}om(C, G)).
\end{equation}

\begin{Example}{Singuluar Cochain Complex}
.  Let $C_{\bullet}$ be:
\begin{equation}
  \begin{tikzcd}
    \dots  \ar[r] & 0 \ar[r] & \Z \ar[r, "0"] & \Z \ar[r, "\cdot 2"] & \Z \ar[r, "0"] & \Z \ar[r] & 0.
  \end{tikzcd}
\end{equation}

Then we have:
\begin{equation}
  H_{*}(C_{\bullet}) =
  \begin{cases}
    \Z & \quad n = 0,\\
    \Z / 2\Z & \quad n = 1,\\
    0 & \quad n = 2,\\
    \Z & \quad n = 3.
  \end{cases}
\end{equation}
If we let $G = \Z$, and dualize, then we obtain:
\begin{equation}
  \begin{tikzcd}
    \dots & \ar[l] 0 & \ar[l] \Z & \ar[l, "0"] \Z & \ar[l, "\cdot 2"] \Z & \ar[l, "0"] \text{\normalfont Hom}(\Z, \Z) \cong \Z & \ar[l] 0.
  \end{tikzcd}
\end{equation}
has cohomology:
\begin{equation}
  H^{*}(C_{\bullet}) =
  \begin{cases}
    \Z & \quad n = 0,\\
    0 & \quad n = 1,\\
    \Z / 2\Z & \quad n = 2,\\
    \Z & \quad n = 3.
  \end{cases}
\end{equation}
These are \textbf{not} the duals of the homology $H_{*}$. Moreover, $G = \Z/2\Z$ has different results. 
\end{Example}

\textbf{\textit{Note: Functoriality.}} If $C_{\bullet} \rightarrow D_{\bullet}$ is a map of cochain complexes, dualizing gives
\begin{align}
  \mathcal{H}om (D_{\bullet}, G) &\rightarrow \mathcal{H}(C_{\bullet}, G) \\
  \varphi &\mapsto \varphi \circ f
\end{align}
which induces a map
\begin{equation}
  H^{*}(\mathcal{H}om(D_{\bullet}, G)) \rightarrow H^{*}(\mathcal{H}om(C_{\bullet}, G)).
\end{equation}

We want to now relate the cohomology of $\mathcal{H}om(C_{\bullet}, G)$ to the homology of $C_{\bullet}$.

\begin{Lemma}{Cohomology and Homology}
. For any chain complex of free Abelian groups, and any Abelian group $G$, there is a natural surjection:
\begin{equation}
  h : H^n(C_{\bullet}; G) \rightarrow \text{\normalfont Hom}(H_n(C_\bullet; G))
\end{equation}
and the sequence
\begin{equation}
  \begin{tikzcd}
    0 \ar[r] & \ker(h) \ar[r] & H^n(C_{\bullet}; G) \ar[r] & \ar[l, dashed, red, bend right=30] \text{\normalfont Hom}(H_{n}(C_{\bullet}; G)) \ar[r] & 0
  \end{tikzcd}
\end{equation}
\coloruwave{red}{splits}.
\end{Lemma}

\begin{proof}
  The proof can be found in my live notes. Please review it in detail.
\end{proof}

\section{Lecture 2}

Our goal for today is to understand $\ker h$.

\begin{Lemma}{``Hom is left exact''}
.  Suppose that
\begin{equation}
  0 \rightarrow A \xrightarrow{f} B \xrightarrow{g} C \rightarrow 0.
\end{equation}
  is a short exact sequence of Abelian groups and $G$ is a group. Then the dual sequence
  \begin{equation}
    0 \rightarrow \text{\normalfont Hom} (C; G) \xrightarrow{g^{*}} \text{\normalfont Hom}(B ; G) \xrightarrow{f^{*}} \text{\normalfont Hom}(A; G)
  \end{equation}
  is exact. If the original sequence was split, then $f^{*}$ is surjective.
\end{Lemma}

\begin{proof}
  See my live notes for the proof. Please review it carefully.
\end{proof}

\textbf{\textit{Remarks.}}

\begin{enumerate}
\item $\text{Hom}(\cdot, G)$ is a contravariant functor, and the property of the lemma is called ``left-exactness''.
\item
  \begin{equation}
    0 \rightarrow \Z \xrightarrow{\cdot n} \Z \rightarrow \Z / n \Z \rightarrow 0
  \end{equation}
  is exact, and $G = \Z$. Dualize:
  \begin{equation}
    0 \rightarrow \text{Hom}(\Z/n\Z, \Z) \xrightarrow{0} \Z \xrightarrow{\cdot n} \Z
  \end{equation}
  is \textit{not} surjective. So, Hom in general is not exact.
\end{enumerate}

Now an argument is developed, regarding degree shifts in cochain complexes, short exact sequences are related via the snake lemma, and we make a statement regarding the cokernel, which measures something interesting. See live notes for details.

In fact: $\text{coker} (i_{n-1}^{*})$ measures the ``non-right-exactness'' of Hom applied to a specific SES:
\begin{equation}\label{eq:sesForCokerNonRightExactness}
  0 \rightarrow B_{n-1} \xrightarrow{i_{n-1}} Z_{n - 1} \rightarrow H_{n - 1}(C_{\bullet}) \rightarrow 0.
\end{equation}

\begin{Corollary}{Cohomology and Homology Homomorphisms}
.  If $H_{n-1}(C_{\bullet})$ is free Abelian, then
\begin{equation}
  H^n(C_{\bullet}, G) \cong \text{\normalfont Hom} (H_{n}(C_{\bullet}, G)).
\end{equation}
\end{Corollary}

\begin{proof}
  If $H_{n-1}(C_{\bullet})$ is free Abelian, then~(\ref{eq:sesForCokerNonRightExactness}) splits. Then apply Hom, get a SES, then the $\text{coker}(i_{n-1}^{*}) = 0$.
\end{proof}

\begin{Definition}{Free Resolution}
.  Let $H$ be an Abelian group. A \textbf{free resolution}\index{free resolution} of $H$ is an exact sequence
\begin{equation}
  \cdots \rightarrow F_2 \rightarrow F_1 \rightarrow F_0 \rightarrow H \rightarrow 0
\end{equation}
where all $F$ are free Abelian.
\end{Definition}

\begin{Lemma}{Free Resolutions of Length 2}
.  Any Abelian group $H$ has a free resolution (of length 2).
\end{Lemma}

\begin{proof}
  $f_0 : \bigoplus_{H\setminus e}\Z \xrightarrow{\pi} H$ a homomorphism mapping $1 \in \Z$ to $h$ induced by $h$. Then
  \begin{equation}
    0 \rightarrow \ker(\pi) \rightarrow F_0 \xrightarrow{\pi} H.
  \end{equation}
\end{proof}

\begin{Corollary}{Cohomology Independent of Free Resolution}
  Suppose that
  \begin{equation}
    \cdots \rightarrow F_2 \rightarrow F_1 \rightarrow F_0 \rightarrow H \rightarrow 0
  \end{equation}
  is a free resolution of $H$, $G$ is any Abelian group. Consider the cochain complex
  \begin{equation}
    0 \rightarrow  \text{\normalfont Hom}(H, G) \xrightarrow{f_0^{*}}\text{\normalfont Hom}(F_0, G) \xrightarrow{f_1^{*}} \text{\normalfont Hom}(F_1, G) \rightarrow \cdots.
  \end{equation}
  The cohomology of this is independent of the free resolution. We call the first cohomology group
  \begin{equation}
    \ker f_2^{*} / \text{im} f_1^{*} = \text{\normalfont Ext}^1(H, G).
  \end{equation}
  $\text{\normalfont Ext}^1$ is the \textbf{derived functor}\index{derived functor} of Hom.
\end{Corollary}

There is also a technical corollary, which I leave in the live notes, since the diagram is complicated.

\begin{Corollary}{Technical Corollary: Chain Morphisms}
.  See live notes.
\end{Corollary}

\printindex

\end{document}
